---
title: "Marxan in R"
subtitle: "Tutorial for ESM 270 Conservation Planning"
author: "Cameryn Brock"
date: "6/27/2020"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

# Introduction

Designing reserve networks requires considering both ecological and socioeconomic factors. Marxan is a commonly used spatial planning tool that quantifies these factors to generate efficient reserve networks. First we will run through a simple example analysis, and then we will use species and parcel data from a 2014-15 Bren Group Project in the Morro Bay Watershed to examine the optimate regional reserve network.

### Priotitizr

`Prioritizr` is an R package that, like Marxan, can guide systematic reserve design by solving conservation prioritization problems. In this tutorial, we will be using the `marxan_problem()` function which can read Marxan input data. More information on the `prioritizr` package can be found at https://mran.microsoft.com/snapshot/2018-03-04/web/packages/prioritizr/vignettes/prioritizr_basics.html. 

A really useful tutorial in using prioritizr, specifically in using summed solutions, can be found here: https://prioritizr.net/articles/tasmania.html. This resource helped frame this tutorial. 

### Gurobi

For the `solve()` function, you must have a solver installed. For the the process with summing and multiple runs process, this must be `gurobi`. This requires following the instructions below.  

1. Download gurobi (https://www.gurobi.com/downloads/gurobi-optimizer-eula/)
2. Request an academic license (https://www.gurobi.com/downloads/end-user-license-agreement-academic/)
2. Verify license in terminal by copying and pasting information in your account > your licenses
3. Must install r package with`install.packages('/Library/gurobi902/mac64/R/gurobi_9.0-2_R_3.6.1.tgz', repos=NULL)`
4. must install slam package with `install.packages("slam")
5. add gurobi and slam libraries

# Getting Started

## Attach Packages

```{r}
library(here)
library(janitor)
library(prioritizr) # marxan
library(sf) # spatial features
library(slam) # to use gurobi
library(gurobi) # solver
library(ggmap) # basemaps
library(patchwork) 
library(tidyverse)
```

If you do not have any of these packages installed, run the following in the console: 
`install.packages("package name")`

Note: this will not work for installing `gurobi`. Instructions are in the introduction section. 

## Example Analysis

Familiarize yourself with the format of Marxan input files (all are text files that can be opened with a text editor) using this small example.

**Data**

* *spec.dat*: species file that contains 24 elements with varying targets expressed as the number of sites protected, constant species penalty factor, and a minimum of 2 occurrences. 

* *pudata.dat*: planning unit file that lists 99 sites with variable site costs. 

* *PUvCVr.dat*: planning unit vs. species file that lists the amount of each species in each planning unit in a codensed format. 

* *bound.dat*: boundary length file that provides the 'effective length' of shared boundary between pairs of planning units. 

### Read in Data

Read in data and select necessary columns for `marxan_problem()`. Find more information with `?marxan_problem()`.

```{r}
spec_samp <- read_tsv(here("sample-data", "spec.dat")) %>% 
  rename("amount" = "target")
# with prioritizr, target is called "prop" (relative) or "amount" (absolute)

pu_samp <- read_tsv(here("sample-data","pudata.dat"))

puvsp_samp <- read_tsv(here("sample-data", "PUvCVr.dat"))

bound_samp <- read_tsv(here("sample-data", "bound.dat"))
```

### Run Marxan

The function `marxan_problem()` in the `prioritizr` package gives us the "canned" approach that works for our purposes. If more customizations are desired, feel free to explore the `problem()` function instead. 

Arguments for `marxan_problem` (for more information, see `?marxan_problem`):

* x: planning units to use in reserve design and their cost (`x = pu_samp`)
* spec: conservation features. Must contain columns for id, name, and prop/amount (`spec = spec_samp`)
* puvspr: the amount of each feature in each planning unit. Must contain columns for pu, species, and amount. (`puvspr = puvsp_samp`)
* bound: boundary data (`bound = bound_samp`)
* blm: boundary length modifier. (`blm = 0.2`)

Also note that in solver, we will use 100 runs (`number_solutions = 100`)

```{r}
marx_samp <- marxan_problem(x = pu_samp,
                              spec = spec_samp,
                              puvspr = puvsp_samp,
                              bound = bound_samp,
                              blm = 0.2)

marx_prob_samp <- marx_samp %>% 
  add_gurobi_solver(gap = 0.15) %>% 
  add_pool_portfolio(method = 2,
                     number_solutions = 10) %>% 
  add_absolute_targets(2)

# gap = optimality gap. 0.15 gives us results closest to that of using marxan software. Learn more under ?add_gurobi_solver 
# add_absolute_targets sets targets for the actual value of features in the study area that need to be represented in the prioritization. Learn more under ?add_absolute_targets. Other option is add_relative_targets(). 

marx_samp_soln <- solve(marx_prob_samp)
```


# Morro Bay Watershed

We will be using a case study (and the data they prepared) from a past group project on the Morro Bay National Estuary Program.

**Data**

*Species data*

* MorroBay_species_polyons: shapefiles of species locations 
* MorroBay_species_pts: point data of species locations 
* MorroBay_habitats: shapefile of location of key habitats
* morro-bay-spec.csv: information on species/habitat ID (the conservation feature), target value for that species (these values are 30% of the # of parcels containing that species or 10% for habitats), and the species name. The 'spf' column is the 'species penalty factor' and is a multiplier that determines the penalty if the conservation objective is *not* met
* spec-name-status.csv: information on species status

*Parcel data*

* morro-bay-pu.csv: includes information on the planning unit ID, the cost of that planning unit, and each parcel's status (value 0 = available for prioritization, value 3 = locking in,  3 = locked out)
* morro-bay-puvspr.csv: includes information on the amount (presence/absence) of each species in each planning unit (pu)

## Analyze All Species

### Read in Data

#### Species and pu information

```{r}
spec <- read_csv(here("morro-bay-data", "morro-bay-spec.csv")) %>% 
  head(140) %>%  # mine reads in extra blank rows - skip if yours does not 
  rename("amount" = "target") # target is called "prop" (relative) or "amount" (absolute)

pu <- read_csv(here("morro-bay-data", "morro-bay-pu.csv")) %>% 
  select(1:3) # mine reads in an extra blank column - skip if yours does not

puvsp <- read_csv(here("morro-bay-data", "morro-bay-puvspr.csv")) %>% 
  select(1:3) %>% 
  head(11849)

status <- read_csv(here("morro-bay-data", "spec-name-status.csv")) %>% 
  select(1:3) %>% 
  head(140)
```

#### Polygons

```{r}
parcels <- read_sf(dsn = here("morro-bay-data"), layer = "MorroBay_parcels") %>% 
  clean_names()

ggplot(data = parcels) +
  geom_sf()
```

### Run Marxan

The function `marxan_problem()` in the `prioritizr` package gives us the "canned" approach that works for our purposes. If more customizations are desired, feel free to explore the `problem()` function instead. 

Arguments for `marxan_problem` (for more information, see `?marxan_problem`):

* x: planning units to use in reserve design and their cost (`x = pu`)
* spec: conservation features. Must contain columns for id, name, and prop/amount (`spec = spec`)
* puvspr: the amount of each feature in each planning unit. Must contain columns for pu, species, and amount. (`puvspr = puvsp`)
* bound: boundary data (we will not use this; `bound = NULL`)
* blm: boundary length modifier. (`blm = 0`)

Also note that in solver, we will use 100 runs (`number_solutions = 100`)

```{r}
marx_mb <- marxan_problem(x = pu,
                         spec = spec, 
                         puvspr = puvsp, 
                         bound = NULL,
                         blm = 0)

marx_mb_problem <- marx_mb %>% 
  add_gurobi_solver(gap = 0.15) %>% 
  add_pool_portfolio(method = 2, number_solutions = 100) 
# gap = optimality gap. 0.15 gives us results closest to that of using marxan software. Learn more under ?add_gurobi_solver 
# method 2 finds a specified number of solutions that are nearest to optimality. Learn more under ?add_pool_portfolio

marx_mb_soln <- solve(marx_mb_problem)
```

### Sum Solutions

```{r}
marx_mb_ssoln <- marx_mb_soln %>% 
  mutate(sum = rowSums(.[6:105])) %>% 
  select(id, cost, status, sum)

hist(marx_mb_ssoln$sum,
     main = "Selection Frequencies",
     xlab = "Number of runs that the unit was selected")
```

### Visualize Results

#### Join polygons with output

```{r}
ssoln_parcels <- inner_join(parcels, marx_mb_ssoln, by = "id")

ggplot(data = ssoln_parcels) +
  geom_sf(aes(fill = sum),
          color = "white",
          size = 0.05) + 
  scale_fill_gradient(low = "slategray2",
                      high = "navy") + 
  labs(fill = "Summed \nSolution") +
  theme_minimal()
```

#### Add basemap

This step is optional, but will make your map look better and is good to learn for further spatial analyses in R. Unfortunately, the options for basemaps in R are limited. Here, I have used the `ggmaps` package because it is the most accessible, however, it is still not very accessible compared to other R packages because you'll need to get an API key. Instructions are here: https://cran.r-project.org/web/packages/ggmap/readme/README.html

For a ggmap cheatsheet, including the different basemaps in ggmaps, see: https://www.nceas.ucsb.edu/sites/default/files/2020-04/ggmapCheatsheet.pdf


```{r}
morrobay <- get_map(location = c(lon = -120.7665, lat = 35.335),
                    zoom = 12,
                    maptype = "terrain-background", # background means no references, omit if want references
                    source = "google")

all_spec <- 
  ggmap(morrobay) +
    geom_sf(data = ssoln_parcels,
            aes(fill = sum),
          color = "white",
          size = 0.1,
          alpha = 0.85,
          inherit.aes = FALSE) + 
  coord_sf(crs = st_crs(4326)) +
  scale_fill_gradient(low = "slategray2",
                    high = "navy") +
  labs(title = "All Species",
       fill = "Summed \nSolution",
       x = NULL,
       y = NULL) +
  theme_minimal()

all_spec
```

## Analyze Endangered Species (Optional)

### Create Dataframes

Here we create spec and puvsp data frames similar to those we had for all species, but containing only endangered or threatened species.

```{r}
end_status <- status %>% 
  mutate("endangered" = case_when(
    str_detect(status, pattern = "endangered") == TRUE ~ "yes",
    str_detect(status, pattern = "threatened") == TRUE ~ "yes",
    T ~ "no")) %>% 
  filter(endangered == "yes")

end_spec <- merge(end_status, spec, by = "id") %>% 
  select(id, amount, spf, name.x) %>% 
  rename("name" = "name.x")

puvsp_id <- puvsp %>% 
  rename("id" = "species")

end_puvsp <- merge(end_status, puvsp_id, by = "id") %>% 
  rename("species" = "id")
```

### Run Marxan

```{r}
marx_end <- marxan_problem(x = pu,
                         spec = end_spec, 
                         puvspr = end_puvsp, 
                         bound = NULL,
                         blm = 0)

marx_end_problem <- marx_end %>% 
  add_gurobi_solver(gap = 0.15) %>% 
  add_pool_portfolio(method = 2, number_solutions = 100)

marx_end_soln <- solve(marx_end_problem)
```

### Sum solutions

```{r}
marx_end_ssoln <- marx_end_soln %>% 
  mutate(sum = rowSums(.[6:105])) %>% 
  select(id, cost, status, locked_in, locked_out, sum)

hist(marx_end_ssoln$sum,
     main = "Selection frequencies",
     xlab = "Number of runs that the units were selected")
```

### Visualize Results

#### Join polygons with output

```{r}
parcels_marx_end <- inner_join(parcels, marx_end_ssoln, by = "id")

ggplot(data = parcels_marx_end) +
  geom_sf(aes(fill = sum),
          color = "white",
          size = 0.05) + 
  scale_fill_gradient(low = "slategray2",
                      high = "navy") + 
  labs(fill = "Summed \nSolution") +
  theme_minimal()
```

#### Add basemap

```{r}
end_spec <- 
  ggmap(morrobay) +
    geom_sf(data = parcels_marx_end,
            aes(fill = sum),
          color = "white",
          size = 0.1,
          alpha = 0.85,
          inherit.aes = FALSE) + 
  coord_sf(crs = st_crs(4326)) +
  scale_fill_gradient(low = "slategray2",
                      high = "navy") + 
  labs(title = "Endangered & Threatened Species",
       fill = "Summed \nSolution",
       x = NULL,
       y = NULL) +
  theme_minimal()

end_spec
```

### Combine Maps with Patchwork

This step isn't necessary, but is really convenient if you want to present your maps together in your report. `patchwork` uses PEMDAS to combine ggplots together into one graphic. For example, `plot_1 + plot_2` will make an image of plot_1 *next to* plot_2, whereas `plot_1 / plot_2` will make an image of plot_1 *over* plot_2. For more information, see https://github.com/thomasp85/patchwork.

First, make the all_spec graph without a legend so that the combined image has only one legend.

```{r}
all_no_lgnd <- all_spec +
  theme(legend.position = "none")
```

Next, combine the graphs with `patchwork`

```{r, fig.width = 9}
spec_graphs <- all_no_lgnd + end_spec

spec_graphs
```

Finally, save the image to add to your report!

```{r}
ggsave("spec-graphs.png")
```

# Questions

The following are some questions to consider in your analysis and report. 

* Are the same management units commonly included in replicate runs? You can check this by looking at the distribution of the summed solutions. What does this pattern suggest? 

* Are your reserves clustered? Why or why not? 

* What was the conservation goal of this preliminary analysis? Is this a *reasonable* goal? How may your results have changed with constraints on your conservation goal or different conservation goals altogether? 

* If you did the extra analysis, how or why would sequencing of parcel protection affect future priorities? 
