---
title: "Marxan in R"
subtitle: "ESM 270 Tutorial"
author: "Cameryn Brock"
date: "6/27/2020"
output: 
  html_document:
    toc: true
    toc_float: true
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

# Introduction

## Marxan



## Priotitizr

Information on the `prioritizr` package can be found at https://mran.microsoft.com/snapshot/2018-03-04/web/packages/prioritizr/vignettes/quick_start.html. 

A really useful example for summed solutions is under "Selection Frequencies" here: https://cran.r-project.org/web/packages/prioritizr/vignettes/tasmania.html 

## Gurobi

For the `solve()` function, you must have a solver installed. For the the process with summing and multiple runs process, this must be `gurobi`. This requires following the instructions below.  

1. Download gurobi (https://www.gurobi.com/downloads/gurobi-optimizer-eula/)
2. Request an academic license (https://www.gurobi.com/downloads/end-user-license-agreement-academic/)
2. Verify license in terminal by copying and pasting information in your acount > your licenses
3. Must install r package with`install.packages('/Library/gurobi902/mac64/R/gurobi_9.0-2_R_3.6.1.tgz', repos=NULL)` (if on a mac and downloaded the most recent version of gurobi, don't download newest version of R! (4.0.0)).
4. must install slam package with `install.packages("slam")
5. add gurobi and slam libraries

# Getting started

## Attach packages

```{r}
library(here)
library(janitor)
library(prioritizr) # marxan
library(sf) # spatial features
library(slam) # to use gurobi
library(gurobi) # solver
library(ggmap) # basemaps
library(patchwork) 
library(tidyverse)
```

If you do not have any of these packages installed, run the following in the console: 
`install.packages("package name")`

Note: this will not work for installing `gurobi`. Instructions are in the introduction section. 

## Read in data

All .csv files in this tutorial are unaltered from the .xlsx files provided in the R drive. They have only been renamed and saved as csvs. The "data" folder is identical to the MorroBay_data folder provided in the R drive, it has only been renamed. 

### Species and pu data

```{r}
spec <- read_csv("morro-bay-spec.csv") %>% 
  head(140) %>%  # mine reads in extra blank rows - skip if yours does not 
  rename("amount" = "target") # target is called "prop" (relative) or "amount" (absolute)

pu <- read_csv("morro-bay-pu.csv") %>% 
  select(1:3) # mine reads in an extra blank column - skip if yours does not

puvsp <- read_csv("morro-bay-puvspr.csv") %>% 
  select(1:3) %>% 
  head(11849)

status <- read_csv("spec-name-status.csv") %>% 
  select(1:3) %>% 
  head(140)
```

### Polygons

```{r}
parcels <- read_sf(dsn = here("data"), layer = "MorroBay_parcels") %>% 
  clean_names()

ggplot(data = parcels) +
  geom_sf()
```

# Analyze all species

## Run Marxan

The function `marxan_problem()` in the `prioritizr` package gives us the "canned" approach that works for our purposes. If more customizations are desired, feel free to explore the `problem()` function instead. 

```{r}
marxan_1 <- marxan_problem(x = pu,
                         spec = spec, 
                         puvspr = puvsp, 
                         bound = NULL,
                         blm = 0)

marxan_1_problem <- marxan_1 %>% 
  add_gurobi_solver(gap = 0.15) %>% 
  add_pool_portfolio(method = 2, number_solutions = 100) # see method meaning under ?add_pool_portfolio

print(marxan_1_problem)

marxan_1_soln <- solve(marxan_1_problem)
```

## Sum solutions

```{r}
marxan_1_ssoln <- marxan_1_soln %>% 
  mutate(sum = rowSums(.[6:105])) %>% 
  select(id, cost, status, locked_in, locked_out, sum)

hist(marxan_1_ssoln$sum)
```

## Create map

### Join polygons with output

```{r}
parcels_marxan_1 <- inner_join(parcels, marxan_1_ssoln, by = "id")

ggplot(data = parcels_marxan_1) +
  geom_sf(aes(fill = sum),
          color = "white",
          size = 0.05) + 
  scale_fill_binned(low = "slategray2",
                      high = "navy") + 
  labs(fill = "Summed \nSolution") +
  theme_minimal()
```

### Add basemap

This step is optional, but will make your map look better and is good to learn for further spatial analyses in R. Unfortunately, the options for basemaps in R are limited. Here, I have used the `ggmaps` package because it is the most accessible, however, it is still not very accessible compared to other R packages.

Using `ggmap` requires getting an API key, which is a bit of a hassle, but worth it. Instructions are here: https://cran.r-project.org/web/packages/ggmap/readme/README.html

For a ggmap cheatsheet, including the different basemaps in ggmaps, see: https://www.nceas.ucsb.edu/sites/default/files/2020-04/ggmapCheatsheet.pdf


```{r}
morrobay <- get_map(location = c(lon = -120.7665, lat = 35.335),
                    zoom = 12,
                    maptype = "terrain-background", # - background means no references, omit if want references
                    source = "google")

all_spec <- 
  ggmap(morrobay) +
    geom_sf(data = parcels_marxan_1,
            aes(fill = sum),
          color = "white",
          size = 0.1,
          alpha = 0.85,
          inherit.aes = FALSE) + 
  coord_sf(crs = st_crs(4326)) +
  scale_fill_gradient(low = "tan",
                    high = "red3") +
  labs(title = "All Species",
       fill = "Summed \nSolution",
       x = NULL,
       y = NULL) +
  theme_minimal()

all_spec
```

# Analyze endangered species

## Create dataframes

```{r}
end_status <- status %>% 
  mutate("endangered" = case_when(
    str_detect(status, pattern = "endangered") == TRUE ~ "yes",
    str_detect(status, pattern = "threatened") == TRUE ~ "yes",
    T ~ "no")) %>% 
  filter(endangered == "yes")

end_spec <- merge(end_status, spec, by = "id") %>% 
  select(id, amount, spf, name.x) %>% 
  rename("name" = "name.x")

puvsp_id <- puvsp %>% 
  rename("id" = "species")

end_puvsp <- merge(end_status, puvsp_id, by = "id") %>% 
  rename("species" = "id")
```

## Run Marxan

```{r}
marxan_end <- marxan_problem(x = pu,
                         spec = end_spec, 
                         puvspr = end_puvsp, 
                         bound = NULL,
                         blm = 0)

marxan_end_problem <- marxan_end %>% 
  add_gurobi_solver(gap = 0.15) %>% 
  add_pool_portfolio(method = 2, number_solutions = 100)

print(marxan_end_problem)

marxan_end_soln <- solve(marxan_end_problem)
```

## Sum solutions

```{r}
marxan_end_ssoln <- marxan_end_soln %>% 
  mutate(sum = rowSums(.[6:105])) %>% 
  select(id, cost, status, locked_in, locked_out, sum)

hist(marxan_end_ssoln$sum)
```

## Create map

### Join polygons with output

```{r}
parcels_marxan_end <- inner_join(parcels, marxan_end_ssoln, by = "id")

ggplot(data = parcels_marxan_end) +
  geom_sf(aes(fill = sum),
          color = "white",
          size = 0.05) + 
  scale_fill_binned(low = "slategray2",
                      high = "navy") + 
  labs(fill = "Summed \nSolution") +
  theme_minimal()
```

### Add basemap

```{r}
end_spec <- 
  ggmap(morrobay) +
    geom_sf(data = parcels_marxan_end,
            aes(fill = sum),
          color = "white",
          size = 0.1,
          alpha = 0.85,
          inherit.aes = FALSE) + 
  coord_sf(crs = st_crs(4326)) +
  scale_fill_gradient(low = "tan",
                      high = "red3") + 
  labs(title = "Endangered & Threatened Species",
       fill = "Summed \nSolution",
       x = NULL,
       y = NULL) +
  theme_minimal()

end_spec
```

# Combine maps with patchwork

This step isn't necessary, but is great if you want to present your maps together. Patchwork uses PEMDAS to combine ggplots together into one graphic. For example, `plot_1 + plot_2` will make an image of the plots *side by side*, whereas `plot_1 / plot_2` will make an image of plot_1 *over* plot_2. For more information, see https://github.com/thomasp85/patchwork

```{r, fig.width = 9}

# Make the all_spec graph without a legend so that the combined image has only one legend

all_no_lgnd <- all_spec +
  theme(legend.position = "none")
  
# Combine graphs

spec_graphs <- all_no_lgnd + end_spec

spec_graphs

# Save graphs as .png to add to your report

ggsave("spec-graphs.png")
```

